BOLD Parameter Estimation Using Particle Filters

### 2010-07-21 doc/thesis/resultsB.tex
removed lots of weasel words...

% TODO include single?
%Before performing tests on a full image, I the particle filter
%on regions deemed active and inactive by statistical parametric mapping
%(SPM). This served the purpose of adjusting the priors as well as the 
%preprocessing based on real world signals. This was actually done before
%the simulations, and then results were carried back the simulations 
%to check consistency.
%After work adjusting parameters, most importantly the weighting function and the 
%priors, particle filter was applied to every voxel in an FMRI image.
%The results of this large scale analysis was a parameter map which was
%then used to calculate normalized square-root MSE image. 
%
%\section{Single-Voxel Analysis}
%The choice of a prior, as discussed previously, is extremely important. While a
%prior may have the potential to give good results, being a monte-carlo algorithm
%there is the possibility for inconsistencies. Thus, increasing the variance
%of the time-constants may allow additional flexibility, it will also cause
%additional model variance. Before running on a full volume I adjusted the 
%prior to ensure that the same input would give the same output 100 times in a 
%row. While this may seem like a given, with a random drawing of the prior,
%this can be difficult. Case in point, the exact same algorithm run twice
%with the time constants all having standard deviations of $.35$ resulted in two
%very different fits, shown in \autoref{fig:badfit_param1}.
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/badfit_param1}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/goodfit_param1}}
%\caption{The same priors gave rise to both fits.}
%\label{fig:badfit_param1}
%\end{figure}
%
%For this reason, I actually lowered the standard deviations of the time
%constants to prevent over-smoothing. This resulted in more consistent,
%though potentially slightly worse fits, two examples of which are 
%shown in \autoref{fig:param2_var}. 
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2a}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2b}}
%\caption{A poor fit, using the same parameters as }
%\label{fig:param2_var}
%\end{figure}
%
%todo: stats of the 100 fits?
%
%\section{Single Time-Series Simulation}
%
\chapter{Real Data}
\label{sec:RealData}
Modeling the BOLD response is of course not of much use if it is only done in a 
single voxel. Although this algorithm will hopefully lead to more novel methods 
of analysis, the standard use for modeling the BOLD signal is to locate activation.
Activation is essentially defined as areas where the input seems to directly drive
the BOLD response, as opposed to there being intermediate factors controlling it. 
Once areas where the BOLD 
model may be accurately estimated are found, integrating the model will allow for accurate
estimation of the state between measurements, which could then be used for more 
advanced analysis, for instance of areas that are being driven by other brain regions.
It all begins with localizing the first activation regions in the
chain. Therefore this section compares the output of the particle filter 
with conventional SPM.

@@

The results from SPM8 are shown in \autoref{fig:hm_canon_spm}, and the results from 
the particle filter are shown in \autoref{fig:hm_canon_pfilter}. Note again that the
output is actually a different scale. Whereas the levels in the SPM map are T-Scores
indicating the probability that the result cannot be explained from a random signal,
the result of the particle filter is simply a normalized square-root of the mean-squared-
residual. Thus, lower is better, and zero is a perfect fit in the particle filter whereas
infinity is a perfect fit in \autoref{fig:hm_canon_spm}. Note that it would seem that there
are quite a few dispersed areas with decent fits in \autoref{fig:hm_canon_pfilter}. Also,
interestingly the lowest error was just above $.7$. 
For the sake of completeness \autoref{fig:hm_canon_pfilter85} shows the same image but
thresheld at $.85$ rather than $1$. Obviously this image seems like a closer match to
the results of SPM, however, the results of \autoref{fig:hm_canon_pfilter} are interesting.

@@

I chose the second voxel (\autoref{fig:comp2}) because it was active in SPM and it would appear to be in a prime
location to be active in the SPM image (given the results in the surrounding voxels). The fit,
however shows just why the voxel did not have a particularly good error. Although certain peaks 
seem to fit rather well the input doesn't seem to correlate that well with the BOLD response. For
instance 75 seconds in, the stimulus is clearly not present, and so the signal should drop off; yet
it doesn't. This is a good example of a voxel that seems to be reacting to something in addition to
the direct stimulus. Even in what should be a rather smooth time-series in SPM, the BOLD signal
looks very noisy.

>>>>

% TODO include single?
%Before performing tests on a full image, I the particle filter
%on regions deemed active and inactive by statistical parametric mapping
%(SPM). This served the purpose of adjusting the priors as well as the 
%preprocessing based on real world signals. This was actually done before
%the simulations, and then results were carried back the simulations 
%to check consistency.
%After work adjusting parameters, most importantly the weighting function and the 
%priors, particle filter was applied to every voxel in an FMRI image.
%The results of this large scale analysis was a parameter map which was
%then used to calculate normalized square-root MSE image. 
%
%\section{Single-Voxel Analysis}
%The choice of a prior, as discussed previously, is extremely important. While a
%prior may have the potential to give good results, being a monte-carlo algorithm
%there is the possibility for inconsistencies. Thus, increasing the variance
%of the time-constants may allow additional flexibility, it will also cause
%additional model variance. Before running on a full volume I adjusted the 
%prior to ensure that the same input would give the same output 100 times in a 
%row. While this may seem like a given, with a random drawing of the prior,
%this can be difficult. Case in point, the exact same algorithm run twice
%with the time constants all having standard deviations of $.35$ resulted in two
%very different fits, shown in \autoref{fig:badfit_param1}.
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/badfit_param1}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/goodfit_param1}}
%\caption{The same priors gave rise to both fits.}
%\label{fig:badfit_param1}
%\end{figure}
%
%For this reason, I actually lowered the standard deviations of the time
%constants to prevent over-smoothing. This resulted in more consistent,
%though potentially slightly worse fits, two examples of which are 
%shown in \autoref{fig:param2_var}. 
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2a}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2b}}
%\caption{A poor fit, using the same parameters as }
%\label{fig:param2_var}
%\end{figure}
%
%todo: stats of the 100 fits?
%
%\section{Single Time-Series Simulation}
%
\chapter{Real Data}
\label{sec:RealData}
Modeling the BOLD response is of course not of much use if it is only done in a 
single voxel. Although this algorithm will hopefully lead to more novel methods 
of analysis, the standard use for modeling the BOLD signal is to locate activation.
Activation is defined as areas where the input seems to directly drive
the BOLD response, as opposed to there being intermediate factors controlling it. 
Once areas where the BOLD 
model may be accurately estimated are found, integrating the model will allow for accurate
estimation of the state between measurements, which could then be used for more 
advanced analysis, for instance of areas that are being driven by other brain regions.
It all begins with localizing the first activation regions in the
chain. Therefore this section compares the output of the particle filter 
with conventional SPM.

@@

The results from SPM8 are shown in \autoref{fig:hm_canon_spm}, and the results from 
the particle filter are shown in \autoref{fig:hm_canon_pfilter}. Note again that the
output is actually a different scale. Whereas the levels in the SPM map are T-Scores
indicating the probability that the result cannot be explained from a random signal,
the result of the particle filter is simply a normalized square-root of the mean-squared-
residual. Thus, lower is better, and zero is a perfect fit in the particle filter whereas
infinity is a perfect fit in \autoref{fig:hm_canon_spm}. Note that it would seem that there
are quite a few dispersed areas with decent fits in \autoref{fig:hm_canon_pfilter}. Also,
the lowest error was just above $.7$. 
For the sake of completeness \autoref{fig:hm_canon_pfilter85} shows the same image but
thresheld at $.85$ rather than $1$. Obviously this image seems like a closer match to
the results of SPM, however, the results of \autoref{fig:hm_canon_pfilter} are interesting.

@@

I chose the second voxel (\autoref{fig:comp2}) because it was active in SPM and it would appear to be in a prime
location to be active in the SPM image (given the results in the surrounding voxels). The fit,
however shows just why the voxel did not have a particularly good error. Although certain peaks 
seem to fit rather well the input doesn't seem to correlate that well with the BOLD response. For
instance 75 seconds in, the stimulus is not present, and so the signal should drop off; yet
it doesn't. This is a good example of a voxel that seems to be reacting to something in addition to
the direct stimulus. Even in what should be a rather smooth time-series in SPM, the BOLD signal
looks very noisy.
